---
title: "Hello TTSI"
author: "Transition Technologies Science"
date: 2024-09-16
categories: ["Clinical Data Science", "R"]
tags: ["R Markdown", "Clinical Data", "Optimization", "TTSI"]
---
    
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

# Introduction to Clinical Data Science at TTSI

This is a demonstration of how **R Markdown** can be used to generate reports and visualizations relevant to **clinical data science**. At **TTSI**, we specialize in study simulations, sample size estimation, and live anomaly detection. Markdown allows us to format reports that are dynamic and reproducible.

You can embed an R code chunk like this to perform basic data analysis, for example:
    
```{r cars}
# Example using the built-in cars dataset
summary(cars)

# Simulate a simple linear model that could be relevant in clinical data analysis
fit <- lm(dist ~ speed, data = cars)
fit
```

# Including Visualizations

At TTSI, visualizations are essential for interpreting complex data from clinical trials. Below is a sample pie chart (Figure \@ref(fig:pie)) demonstrating a common method for displaying categorical data distributions, like treatment groups in a clinical study.

```{r pie, fig.cap='Distribution of Treatment Groups in a Study.', tidy=FALSE}
par(mar = c(0, 1, 0, 1))
pie(
    c(150, 100, 50),
    c('Treatment A', 'Treatment B', 'Placebo'),
    col = c('#1f77b4', '#ff7f0e', '#2ca02c'),
    init.angle = -30, border = NA
)
```

# Conclusion

This document showcases how **R Markdown** can be effectively utilized at **TTSI** for producing clinical reports, visualizations, and data analysis outputs that are dynamic and reproducible.
