---
title: "This is just testing!"
author: "Kamil Sijko"
date: 2024-09-16
categories: ["Clinical Data Science", "R"]
tags: ["R Markdown", "Clinical Data", "Optimization", "TTSI"]
---
    
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
library(ggplot2)
```

# Introduction to Clinical Data Science at TTSI

This is a demonstration of how **R Markdown** can be used to generate reports and visualizations relevant to **clinical data science**. At **TTSI**, we specialize in study simulations, sample size estimation, and live anomaly detection. Markdown allows us to format reports that are dynamic and reproducible.

You can embed an R code chunk like this to perform basic data analysis, for example:
    
```{r cars}
# Load the ToothGrowth dataset
data(ToothGrowth)

# Inspect the first few rows
head(ToothGrowth)

# Fit a linear model to predict tooth length based on the dose of vitamin C
fit <- lm(len ~ dose, data = ToothGrowth)
fit |> 
  broom::tidy() |> 
  knitr::kable(digits = 2)
```

# Including Visualizations

At TTSI, visualizations are essential for interpreting complex data from clinical trials. Below is a sample pie chart (Figure \@ref(fig:pie)) demonstrating a common method for displaying categorical data distributions, like treatment groups in a clinical study.

```{r pie}
plot <- 
  ggplot(ToothGrowth, aes(x = dose, y = len)) +
  geom_point() +  # Add points for the raw data
  geom_smooth(method = "lm", col = "blue") +  # Add the linear model line
  labs(title = "Linear Model: Tooth Length vs Vitamin C Dose",
       x = "Dose of Vitamin C (mg)",
       y = "Tooth Length") 
plot
```

# Conclusion

This document showcases how **R Markdown** can be effectively utilized at **TTSI** for producing clinical reports, visualizations, and data analysis outputs that are dynamic and reproducible.
